Synthesizing Quotient Lenses

### 2017-10-18 papers/QuotientLensSynth/main.tex
add new graphs and fix some text

\newif\ifdraft\drafttrue  % set true to show comments
\newif\ifplentyoftime\plentyoftimefalse  % :-)

@@

\usepackage{amsmath, amssymb, verbatim, enumerate, graphicx, centernot, tikz,
array, mathtools, bussproofs, stmaryrd, enumitem, stackengine}
\usepackage{relsize}
\usepackage{listings}
\lstset{
  language=Caml,
  basicstyle=\upshape\sffamily,
  keywordstyle=\upshape\sffamily\color{dkblue},
  keepspaces=true,
  framexleftmargin=1ex,
  framexrightmargin=1ex,
  showstringspaces=false,
  commentstyle=\itshape\rmfamily,
  columns=flexible,
  % BCP: I find this distracting:
  % stringstyle=\sffamily\color{dkred},
  mathescape
}
\newcommand{\cd}[1]{\lstinline[backgroundcolor=\color{white}]$#1$}

@@

\newcommand{\OpticianRuntime}{OO}
\newcommand{\SystemOnOptician}{SO}
\newcommand{\SystemOnBenchmarks}{SS}

@@

{\em Lenses}~\cite{Lenses} specify bidirectional 
transformations between different data formats.  Writing a single lens
rather than two separate uni-directional transformations saves both
keystrokes and bugs, by reducing redundancy and by automatically
guaranteeing strong ``lens laws'' ensuring that the forward and backward
transformations interact well.  
%
{\em Quotient lenses}~\cite{quotientlenses} are a generalization of lenses
where each of the data formats is quotiented by some equivalence relation
and the lens laws are taken modulo equivalence.  This allows lenses to be
both simpler and more flexible, since whole equivalence classes of data
values can be handled all at once and inessential details like whitespace
and ordering of fields can be ignored.

\smallskip

However, while plain lenses are widely used in both standalone systems and
popular libraries, Boomerang~\cite{Boomerang} is the only implementation
currently offering quotient lenses.  One reason for this is that there is
some annoying redundancy in the quotient lens notation: The programmer must
separately specify (1) the data formats on each side of the lens, (2) the
equivalence relations over them, and (3) a lens between the canonical
representatives of the equivalence classes, even though, typically, these
three share a great deal of structure.  Moreover, writing the lens between
the equivalence classes can be challenging due to the rather fiddly, mainly
due to ``non-ambiguity'' conditions in the lens well-formedness rules.

\smallskip

In this paper, we address both of these issues.  First, we propose a novel
language of {\em Quotient Regular Expressions}, or QREs, which allows
programmers to specify a broad class of equivalence relations over regular
languages using a compact and convenient notation in which the regular
language and its equivalence classes are captured by a single expression.
We then define {\em QRE lenses}, a class of bijective quotient lenses that
are typed using QREs, and show how to {\em synthesize} QRE lenses from their
QRE types and a set of input-output examples.  We demonstrate the
practicality of this approach by synthesizing QRE lenses between several
real-world data formats from the {\tt data.gov} database.
\end{abstract}

\section{Introduction}
\ifplentyoftime
\bcp{Not sure there is time to fix this, but the intro seems very long and
  technical---it takes a long time to get to saying what our actual
  contributions are!  Wonder if we should move most of it to a second
  ``Example'' section.} 
\fi
{\em Bidirectional programming languages} such as
Boomerang~\cite{boomerang}, Augeas~\cite{augeas}, XSugar~\cite{xsugar},
biXid~\cite{bixid} and X/Inv~\cite{xinv}\bcp{there are many others; we
  should mention all of them} enable a programmer to express a
transformation that maps some input data to some output data, as well as a
transformation that maps the output data back into the input data in a single
expression. An example of a {\em bidirectional transformation} which we shall
revisit severally in this paper is a program which converts citation records
such as
\begin{verbatim}
@Book {conway,
  Author = "J. H. Conway",
  Title = {Regular Algebra and Finite Machines}
}
\end{verbatim}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{attheedges}
  \caption{Quotient Lens with ``Canonizers at the Edges''\sam{This is a
      screenshot from Nate's thesis. Are we allowed to use it?}\bcp{We can
      (if we put ``from [Nate's thesis]'' in the caption), but I don't think
      we want the full complexity of asymmetric lenses here.  We should draw
      the figure again.  (I have the OmniGraffle file for it, so this is
      easy if I know how it should look.)}}
  \label{fig:attheedges}
\end{figure}

@@

In summary, our main contributions are:
\begin{enumerate}
\item We introduce a novel language of {\em Quotient Regular Expressions}
(QREs) that provide compact, convenient notation for a natural and useful
class of equivalence relations on regular languages (Section~\ref{QRE}).
\item We define a language of {\em QRE lenses}, a subset of bijective quotient
lenses whose types are given by QREs.  Our main technical contribution is a
normal form for QRE lenses and a proof that normal forms are
closed under left quotienting, right quotienting, composition, and regular
operators (Section~\ref{QRE-lenses}).
\item Using this normal form, we reduce the problem of {\em synthesizing}
QRE lenses from their QRE types and a set of examples to the previously studied
problem of synthesizing bijective lenses between regular languages 
(Section~\ref{synth}).
\item We extend the Boomerang {\em implementation} with QREs, QRE lens operators,
and QRE lens synthesis and demonstrate its applicability by using it to
synthesize QRE lenses between several real-world data formats from the
{\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

@@

\subsection{Overview}

@@

      q & W(q) & K(q) \\ \hline
      R & R & R \\
      R \mapsto s & W(q) & s \\
      \squash{R}{R'}{f} & R \sep R' & R' \\
      \normalize{R}{R'}{f} & R & R' \\
      q_1 \circ  q_2 & W(q_2) & K(q_1) \\
      q_1 \cdot q_2 & W(q_1) \cdot W(q_2) & K(q_1) \cdot K(q_2) \\
      q_1 \sep q_2 & W(q_1) \sep W(q_2) & K(q_1) \sep K(q_2) \\
      q^* & W(q)^* & K(q)^* \\
    \end{array}
  \]
\[
\begin{array}{r@{\quad}l}
W( \perm{q_1, \ldots, q_n}{q} ) = &
\bigcup \limits_{\sigma \in S_n} W(q_{\sigma(1)}) \cdot W(q) \cdot \ldots \cdot
W(q) \cdot W(q_{\sigma(n)})
\\
K( \perm{q_1, \ldots, q_n}{q} ) = &
 K(q_1) \cdot K(q) \cdot \ldots \cdot K(q) \cdot K(q_n) 
\end{array}
\]
  \caption{Whole and Kernel Regular Expressions\bcp{try putting the 
      last two clauses into the same format as the
      others (using $\mathit{qq} = ...$ and maybe reducing spacing around
      cdots, etc.)}}
  \label{fig:wk}
\end{figure}

@@

      \begin{prooftree}
\AxiomC{$q_1 : c_1 \Leftrightarrow d_1 $}
\AxiomC{$q_2 : c_2 \Leftrightarrow d_2$}
\AxiomC{$\mathcal{L}(W(c_1)) \cap \mathcal{L}(W(c_2)) = \varnothing$}
\AxiomC{$\mathcal{L}(W(d_1)) \cap \mathcal{L}(W(d_2)) = \varnothing$}
\QuaternaryInfC{$q_1 \sep q_2: (c_1 \sep c_2)
\Leftrightarrow (d_1 \sep d_2)$}
\end{prooftree}
  $$
  \llbracket q_1 \sep q_2 \rrbracket.get(s) = 
  \begin{cases}
  \llbracket q_1 \rrbracket.get (s) & \text{if } s \in \mathcal{L}(W(c_1))\\
  \llbracket q_2 \rrbracket.get (s) & \text{if } s \in \mathcal{L}(W(c_2))\\
  \end{cases}$$
  $$\llbracket q_1 \sep q_2 \rrbracket.put(s) = 
  \begin{cases}
  \llbracket q_1 \rrbracket.put (s) & \text{if } s \in \mathcal{L}(W(d_1))\\
  \llbracket q_2 \rrbracket.put (s) & \text{if } s \in \mathcal{L}(W(d_2))\\
  \end{cases}
  $$
  \caption{Denotation and Typing Rules for QRE Lenses\bcp{I think we can
      display this more compactly if we make the inference rules narrower by
    stacking the premises vertically and then put each inference rule {\em
      beside} its semantics definition.}}
  \label{fig:qlenssemantics}
\end{figure}

@@

We have integrated QREs and quotient lens synthesis into Boomerang with an
additional \afm{loc} lines of OCaml code.

All evaluations were performed on a 2.5 GHz Intel Core i7 processor with 16 GB
of 1600 MHz DDR3 running macOS Sierra.

@@

\subsection{Benchmark Suite Construction} 
Our benchmarks were constructed from two sources:
\begin{enumerate}
\item \afm{39} of our benchmarks were derived from the Optician system.  These
  benchmarks were a combination of custom benchmarks, benchmarks derived from
  FlashFill, and benchmarks derived from Augeas.  Many of the formats mapped
  between had to be modified to work with the bijectivity constraints
  Optician required.  Commonly, one format had whitespace where the other did
  not allow for whitespace.  To address the bijectivity issues surrounding this,
  we added whitespace to the format that typically did not have whitespace.  We
  were able to remove these alterations in adopting these benchmarks to our
  system.

\item \afm{8} of our benchmarks were derived from data.gov.  Data.gov contains
  many transformations between formats.  In addition to the formats previously
  adopted for Optician, we synthesized transformers between these formats.
\end{enumerate}

@@

@@

We summarize the results of these experiments in Figure~\ref{?}.  \Name{} was
able to synthesize all of Optician's benchmarks at a speed competitive with
Optician.  There was additional overhead in calculating $W(q)$ and $K(q)$,
resulting in a very slight decrease in performance.

>>>>

\newif\ifdraft\drafttrue  % set true to show comments

@@

\usepackage{amsmath, amssymb, verbatim, enumerate, graphicx, centernot, tikz,
  array, mathtools, bussproofs, stmaryrd, enumitem, stackengine, subcaption}
\captionsetup{compatibility=false}
\usepackage{relsize}
\usepackage{listings}
\lstset{
  language=Caml,
  basicstyle=\upshape\sffamily,
  keywordstyle=\upshape\sffamily\color{dkblue},
  keepspaces=true,
  framexleftmargin=1ex,
  framexrightmargin=1ex,
  showstringspaces=false,
  commentstyle=\itshape\rmfamily,
  columns=flexible,
  stringstyle=\sffamily\color{dkred},
  mathescape
}
\newcommand{\cd}[1]{\lstinline[backgroundcolor=\color{white}]$#1$}

@@

\newcommand{\OpticianRuntime}{\textbf{OO}}
\newcommand{\SystemOnOptician}{\textbf{SO}}
\newcommand{\SystemOnBenchmarks}{\textbf{SS}}

@@

{\em Lenses}~\cite{Lenses} are programs that specify bidirectional
transformations between different data formats.  Writing a single lens
rather than two separate uni-directional transformations saves both
keystrokes and bugs, by reducing redundancy and by automatically
guaranteeing strong ``lens laws'' ensuring that the forward and backward
transformations interact well.
%
{\em Quotient lenses}~\cite{quotientlenses} are a generalization of lenses
where the two data formats are considered modulo some equivalence relation.
The use of quotients allows the programmer to ignore inessential differences
in the data, such as the amount of whitespace or the ordering of fields.

Quotient lenses can be difficult to program, requiring the programmer to
give both a specification for the data on each side of the lens and a
separate specification for the equivalence relations on the data.  The
programmer must also manually implement a lens program that maps between the
formats while respecting the equivalence relations---this is itself a
challenging task thanks to the restrictions on how lenses can be combined
while still satisfying the lens laws.  \bcp{Don't really understand that
  paragraph!}

This paper addresses both of these challenges\bcp{what are the two
  challenges?}. First, we propose the language of {\em Quotient Regular
  Expressions}, or QREs, which define a broad class of equivalence relations
directly on regular languages using a compact and convenient notation. This
eliminates the need to specify the data and the equivalence relation on the
data separately. We then define {\em QRE lenses}, a class of quotient lenses
that are typed using QREs, and show how to {\em synthesize} QRE lenses from
their QRE types and a set of input-output examples. Finally, we demonstrate
the applicability of this approach by synthesizing QRE lenses between
several real-world data formats from the {\tt data.gov} database.
\end{abstract}

\section{Introduction}
\bcp{Not sure there is time to fix this, but the intro seems very long and
  technical---it takes a long time to get to saying what our actual
  contributions are!  Wonder if we should move most of it to a second
  ``Example'' section.} 
{\em Bidirectional programming languages} such as
Boomerang~\cite{boomerang}, Augeas~\cite{augeas}, XSugar~\cite{xsugar},
biXid~\cite{bixid} and X/Inv~\cite{xinv}enable a programmer to express a
transformation that maps some input data to some output data, as well as a
transformation that maps the output data back into the input data in a single
expression. An example of a {\em bidirectional transformation} which we shall
revisit severally in this paper is a program which converts citation records
such as
\begin{verbatim}
@Book {conway,
  Author = "J. H. Conway",
  Title = {Regular Algebra and Finite Machines}
}
\end{verbatim}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{attheedges}
  \caption{Quotient Lens with ``Canonizers at the Edges''\sam{This is a
      screenshot from Nate's thesis. Are we allowed to use it?}\bcp{We can
      (if we put ``from [Nate's thesis]'' in the caption), but I don't think
      we want the full complexity of asymmetric lenses here.  We should draw
      the figure again.  (I might have the OmniGraffle file for it lying
      around someplace...)}}
  \label{fig:attheedges}
\end{figure}

@@

In summary, our main contributions are:
\begin{enumerate}
\item We introduce a novel language of {\em Quotient Regular Expressions}
(QREs) that provide compact, convenient notation for a natural and useful
class of equivalence relations on regular languages (Section~\ref{QRE}).
\item We define a language of {\em QRE lenses}, a form of quotient lenses
whose types are given by QREs.  Our main technical contribution is a normal
form for QRE lenses and a proof that QRE lenses in normal form are closed
under left quotienting, right quotienting, composition, and regular
operators (Section~\ref{QRE-lenses}).
\item Using this normal form, we reduce the problem of {\em synthesizing}
QRE lenses from their QRE types and a set of examples to the previously studied
problem of synthesizing bijective lenses between regular languages 
(Section~\ref{synth}).
\item We extend the Boomerang {\em implementation} with QREs, QRE lens operators,
and QRE lens synthesis and demonstrate its applicability by using it to
synthesize QRE lenses between several real-world data formats from the
{\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

@@

@@

      q & W(q) & K(q) \\ \hline
      R & R & R \\
      R \mapsto s & W(q) & s \\
      \squash{R}{R'}{f} & R \sep R' & R' \\
      \normalize{R}{R'}{f} & R & R' \\
      q_1 \circ  q_2 & W(q_2) & K(q_1) \\
      q_1 \cdot q_2 & W(q_1) \cdot W(q_2) & K(q_1) \cdot K(q_2) \\
      q_1 \sep q_2 & W(q_1) \sep W(q_2) & K(q_1) \sep K(q_2) \\
      q^* & W(q)^* & K(q)^* \\
    \end{array}
  \]
\[
\begin{array}{r@{\quad}l}
W( \perm{q_1, \ldots, q_n}{q} ) = &
\bigcup \limits_{\sigma \in S_n} W(q_{\sigma(1)}) \cdot W(q) \cdot \ldots \cdot
W(q) \cdot W(q_{\sigma(n)})
\\
K( \perm{q_1, \ldots, q_n}{q} ) = &
 K(q_1) \cdot K(q) \cdot \ldots \cdot K(q) \cdot K(q_n) 
\end{array}
\]
  \caption{Whole and Kernel Regular Expressions\bcp{left-justify columns,
      and try putting the last two clauses into the same format as the
      others (using $\mathit{qq} = ...$ and maybe reducing spacing around
      cdots, etc.)}}
  \label{fig:wk}
\end{figure}

@@

      \begin{prooftree}
\AxiomC{$q_1 : c_1 \Leftrightarrow d_1 $}
\AxiomC{$q_2 : c_2 \Leftrightarrow d_2$}
\AxiomC{$\mathcal{L}(W(c_1)) \cap \mathcal{L}(W(c_2)) = \varnothing$}
\AxiomC{$\mathcal{L}(W(d_1)) \cap \mathcal{L}(W(d_2)) = \varnothing$}
\QuaternaryInfC{$q_1 \sep q_2: (c_1 \sep c_2)
\Leftrightarrow (d_1 \sep d_2)$}
\end{prooftree}
  $$
  \llbracket q_1 \sep q_2 \rrbracket.get(s) = 
  \begin{cases}
  \llbracket q_1 \rrbracket.get (s) & \text{if } s \in \mathcal{L}(W(c_1))\\
  \llbracket q_2 \rrbracket.get (s) & \text{if } s \in \mathcal{L}(W(c_2))\\
  \end{cases}$$
  $$\llbracket q_1 \sep q_2 \rrbracket.put(s) = 
  \begin{cases}
  \llbracket q_1 \rrbracket.put (s) & \text{if } s \in \mathcal{L}(W(d_1))\\
  \llbracket q_2 \rrbracket.put (s) & \text{if } s \in \mathcal{L}(W(d_2))\\
  \end{cases}
  $$
  \caption{Denotation and Typing Rules for QRE Lenses}
  \label{fig:qlenssemantics}
\end{figure}

@@

We have integrated QREs and quotient lens synthesis into Boomerang allowing QREs
and synthesis tasks to be expressed alongside existing Boomerang expressions.

All evaluations were performed on a 2.5 GHz Intel Core i7 processor with 16 GB
of 1600 MHz DDR3 running macOS High Sierra.

@@

\subsection{Benchmark Suite Construction} 
Our benchmarks were constructed from two sources:
\begin{enumerate}
\item 39 of our benchmarks were derived from the Optician system.  These
  benchmarks were a combination of custom benchmarks, benchmarks derived from
  FlashFill, and benchmarks derived from Augeas.  Many of the formats mapped
  between had to be modified to work with the bijectivity constraints
  Optician required.  Commonly, one format had whitespace where the other did
  not allow for whitespace.  To address the bijectivity issues surrounding this,
  we added whitespace to the format that typically did not have whitespace.  We
  were able to remove these alterations in adopting these benchmarks to our
  system.

\item 8 of our benchmarks were derived from data.gov.  Data.gov contains
  many transformations between formats.  In addition to the formats previously
  adopted for Optician, we synthesized transformers between these formats.
\end{enumerate}

@@

\begin{figure}
  \centering
  \begin{subfigure}[b]{.49\textwidth}
    \centering
    \includegraphics{generated-graphs/times_opt}
    \caption{}
    \label{subfig:lenssize}
  \end{subfigure}
  \begin{subfigure}[b]{.49\textwidth}
    \includegraphics{generated-graphs/times_new}
    \caption{}
    \label{subfig:examplesused}
  \end{subfigure}
  \caption{Runtimes of Optician and \Name{}.
    In (a), we run Optician and \Name{} on Optician's benchmarks.  We find that
    there is only a negligible difference incurred by Optometrist's overhead.
    In (b), we run Optometrist on its benchmark suite.  We find that it is able
    to synthesize all quotient lenses in under 20 seconds, and typically
    finished in under 5 seconds. }
  \label{fig:times}
\end{figure}

@@

We summarize the results of these experiments in Figure~\ref{fig:times}.
\Name{} was
able to synthesize all of Optician's benchmarks at a speed competitive with
Optician.  There was additional overhead in calculating $W(q)$ and $K(q)$,
resulting in a very slight decrease in performance.
